---
title: "Data Documentation"
subtitle: "NFL Quarterback Retirement Research"
date: "2026-02-25"
author: "Emma Nguyen"
format: 
  pdf
html-math-method: mathjax
execute:
  eval: TRUE
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
library(here)
```

# Research Overview

**Purpose:** Survival Analysis of Quarterback retirement in National Football League (NFL).

**Built by:** `00_nfl_EDA.Rmd`

-   **Event:** Retirement = last season in NFL (no games after season *t*).
-   **Time:** Seasons since debut (`Career_Time`).
-   **Censoring:** Active through end of data → `Retired = 0`. Data **stops at 2016**; \~98 QBs censored (EDA).

**Raw inputs:** Basic_Stats.csv, Game_Logs_Quarterback.csv, Career_Stats_Passing.csv

-   **Basic_Stats.csv**: Player-level demographics and roster information, including name, player ID, age, height, weight, current status/team (excluded), position, and experience/years played.

-   **Game_Logs_Quarterback.csv**: Game-level logs for quarterbacks only, with season/year, week and game date, home/away and outcome, games played, passing attempts/completions/yards, sacks and sack yards, rushing yards, interceptions, touchdowns, and other game stats.

-   **Career_Stats_Passing.csv**: Season-by-season passing summaries (one row per player-season) that aggregate key passing statistics over each year of a player’s career; used mainly for cross-checking.

**Sources:** [Kaggle NFL Statistics](https://www.kaggle.com/datasets/kendallgillies/nflstatistics/data) (scraped from official [NFL](http://www.nfl.com) website)

# Dataset:

**File:** `5_ALL_DATA.csv`

**Unit:** Each row is one QB (Quarterback). **Scope:** QBs with ≥1 row in Game_Logs_Quarterback. **Merge:** Career + Debut-Year + First-Three-Year Summary Statistics

Note: No raw career totals used as predictors to avoid leaking survival time.

## Load Data and Dimensions

```{r load data}
nfl_data <- read_csv(
  here("nfl", "data", "cleaned", "5_ALL_DATA.csv"),
  show_col_types = FALSE)

nfl_data
# 35 columns, 466 rows (Quarterbacks)
dim(nfl_data)
```

# Variable Reference

### *Identifiers and Outcome*

+-------------+-----------+---------------------------------------------------------------------------+
| Variable    | Type      | Description                                                               |
+=============+===========+===========================================================================+
| `Player_Id` | character | Unique Player ID.                                                         |
+-------------+-----------+---------------------------------------------------------------------------+
| `Name`      | character | Player name.                                                              |
+-------------+-----------+---------------------------------------------------------------------------+
| `Retired`   | 0/1       | **Outcome.** 1 = last season \< 2016; 0 = censored (active through 2016). |
+-------------+-----------+---------------------------------------------------------------------------+

### *Demographics and Career Frame*

+-------------------+-----------+-------------------------------------------------------------------------------+
| Variable          | Type      | Description / formula                                                         |
+===================+===========+===============================================================================+
| `Debut_Age`       | numeric   | Age at debut: `Age - (2016 - First_Year)` (Original Data: Age = age in 2016). |
+-------------------+-----------+-------------------------------------------------------------------------------+
| `Height (inches)` | numeric   | Height of Player in inches.                                                   |
+-------------------+-----------+-------------------------------------------------------------------------------+
| `Weight (lbs)`    | numeric   | Weight of Player in pounds.                                                   |
+-------------------+-----------+-------------------------------------------------------------------------------+
| `First_Year`      | numeric   | Debut year (\~1970–2016).                                                     |
+-------------------+-----------+-------------------------------------------------------------------------------+
| `Career_Time`     | numeric   | **Survival time:** seasons from debut = `Last_Year - First_Year + 1`.         |
|                   |           |                                                                               |
|                   |           | Ex. 11 seasons - 2001 to 2011                                                 |
+-------------------+-----------+-------------------------------------------------------------------------------+

### *Game counts (ONLY Regular-Season)*

| Variable           | Description                     |
|--------------------|---------------------------------|
| `Career_Games`     | Total Games Played in Career.   |
| `Debut_Year_Games` | Total Games in debut season.    |
| `Three_Year_Games` | Total Games in first 3 seasons. |

### *Derived ratios (Career\_\*\| Debut\_\* \| Three_Year\_\*)*

-   **TD_INT:** `TD Passes / max(Ints, 1)` - Touchdown to Interception ratio; Infer accuracy/efficiency; higher → longer career (EDA).
-   **RUN_PASS:** `Rushing Yards / max(Passing Yards, 1)` — mobility; can be \<0 (sacks); higher → more hits, earlier retirement (EDA).

### *Completion and Per-Game stats (\*\_PG)*

-   **Completion_Percentage** (Career / Debut / Three-Year): Number of Throws Completions / Attempts; Infer accuracy/efficiency
-   **Per-Game Statistics:** `*_PG`, `Debut_*_PG`, `Three_Year_*_PG` for Yards (passing), Sacks, TD (touchdowns), Ints (interceptions), Rush (rushing yards).
    -   **Denominator** = Career_Games, Debut_Year_Games, or Three_Year_Games;
        -   **NA** when denominator = 0.

# Missingness and Ranges

**Missingness:** Debut stats have more NAs (when there are no debut games or 0 attempts). Three-year window has fewer NAs.

-   Most missingness is 'Debut_Completion_Percentage' with 181 of 466 rows NA.

```{r NAs}
missingness <- colSums(is.na(nfl_data))
missingness[missingness > 0]
max(missingness) 
```

**Ranges:**

-   First_Year (Debut Year): 1970–2016

-   RUN_PASS can be negative

-   Many zeros in TD_INT / per-game for short careers. Consider filtering e.g. `Big_T > 2` or `Total_Games > 0` for modeling.

# Data Structure and Summary

```{r summary}
glimpse(nfl_data)
summary(nfl_data)
```

# Limitations

+----------------+---------------------------------------------------------------------------------------------------------------------------------------+
| Topic          | Note                                                                                                                                  |
+================+=======================================================================================================================================+
| **Retirement** | Inferred (no games after season *t*). Misclassification possible (injury, etc.).                                                      |
+----------------+---------------------------------------------------------------------------------------------------------------------------------------+
| **Censoring**  | Cutoff 2016; QBs still active if Retired = 0. Consider sensitivity to cutoff.                                                         |
+----------------+---------------------------------------------------------------------------------------------------------------------------------------+
| **Leakage**    | Remove career totals (there for reference); per-game and ratios (debut/3-year windows) used so predictors don't encode career length. |
+----------------+---------------------------------------------------------------------------------------------------------------------------------------+

# Data Pipeline

-   **Script:** `00_nfl_EDA.Rmd` containing full cleaning, formulas, and visuals.
-   **Final Output:** `nfl/data/cleaned/5_ALL_DATA.csv`.
-   **Intermediates Files:** `1_cleaned_nfl_data.csv`, `2_reorder_nfl_data.csv`, `3_cleaned_nfl_debut.csv`, `4_cleaned_nfl_3year.csv`.

# Intermediate NFL Datasets

Inspired by the UCI data documentation, this section briefly describes the **intermediate NFL datasets** that lead up to the final `5_ALL_DATA.csv` file, and how each is constructed from the raw inputs using `00_nfl_EDA.Rmd`.

## 1. Career-level cleaned dataset: `1_cleaned_nfl_data.csv`

-   **File path**: `nfl/data/cleaned/1_cleaned_nfl_data.csv`\
-   **Generating code**: within `nfl/code/00_nfl_EDA.Rmd` (section “Filter Files for Statistics only on Quarterbacks” and “Join Basic_Stats and QB_Logs File”).

**Key steps:**

-   Identify quarterbacks by taking distinct `Player Id` values from `Game_Logs_Quarterback.csv`, then restrict `Basic_Stats.csv` to these IDs to form `player_clean` (keeping age, height, weight, experience, and years played).\
-   Restrict `Game_Logs_Quarterback.csv` to **regular season** games only and compute `qb_career_summary` by grouping by player and summarizing: first and last year played, total seasons and games, and career totals of key stats (passing yards, sacks, interceptions, touchdowns, rushing yards).\
-   Define survival-style quantities: `Time` (seasons played), `Event` (retirement indicator based on whether Last_Year \< 2016), and ratios `TD_INT` and `RUN_PASS`.\
-   Left-join `qb_career_summary` with `player_clean` to add demographics, compute `Debut_Age = Age − (2016 − First_Year)`, and rename `Time` to `Experience.`

```{r message = FALSE, warning = FALSE}
nfl_1 <- read_csv(here("nfl", "data", "cleaned", "1_cleaned_nfl_data.csv"),
                  show_col_types = FALSE)
colnames(nfl_1)
```

## 2. Per-game career dataset: `2_reorder_nfl_data.csv`

-   **File path**: `nfl/data/cleaned/2_reorder_nfl_data.csv`\
-   **Generating code**: `qb_combined3` block in `00_nfl_EDA.Rmd`.

**Key steps:**

-   Starting from `1_cleaned_nfl_data.csv`, create per-game versions of key career totals:\
    `Yards_per_game`, `Sacks_per_game`, `TD_per_game`, `Ints_per_game`, and `Rush_per_game` using `Total_Games` as the denominator.\
-   Drop raw career totals (`Passing Yards`, `Sacks`, `TD Passes`, `Ints`, `Rushing Yards`, and raw attempts/completions) so that analyses focus on per-game rates and ratios rather than totals that are functions of survival time.\
-   Write the result to `2_reorder_nfl_data.csv`, variables are reordered for readability purposes.

## 3. Debut-year dataset: `3_cleaned_nfl_debut.csv`

-   **File path**: `nfl/data/cleaned/3_cleaned_nfl_debut.csv`\
-   **Generating code**: “Debut Year Statistics” section in `00_nfl_EDA.Rmd`.

**Key steps:**

-   From regular-season QB logs, construct `qb_first_year` by keeping only games from each player’s **debut season** (minimum `Year` per `Player Id`).\
-   Summarize debut-season totals for the same core stats (passing yards, sacks, touchdowns, interceptions, rushing yards) and the number of games played in the debut year.\
-   Join with `player_clean` to add height and weight; compute debut-year efficiency measures (`Debut_TD_INT`, `Debut_RUN_PASS`, and `Debut_Completion_Percentage`) and per-game rates (`Debut_Yards_per_game`, `Debut_Sacks_per_game`, `Debut_TD_per_game`, `Debut_Ints_per_game`, `Debut_Rush_per_game`).\
-   Save the resulting debut-only summary as `3_cleaned_nfl_debut.csv`.

```{r message = FALSE, warning = FALSE}
nfl_debut <- read_csv(here("nfl", "data", "cleaned", "3_cleaned_nfl_debut.csv"),
                      show_col_types = FALSE)
colnames(nfl_debut)
```

## 4. First-three-years dataset: `4_cleaned_nfl_3year.csv`

-   **File path**: `nfl/data/cleaned/4_cleaned_nfl_3year.csv`\
-   **Generating code**: “Three Year Statistics” section in `00_nfl_EDA.Rmd`.

**Key steps:**

-   Build `qb_3_years` by keeping only games from a QB’s **first three seasons** (years ≤ debut year + 2).\
-   Summarize totals over those three seasons, then compute three-year efficiency ratios (`Three_Year_TD_INT`, `Three_Year_RUN_PASS`, `Three_Year_Completion_Percentage`) and per-game rates (`Three_Year_Yards_per_game`, `Three_Year_Sacks_per_game`, `Three_Year_TD_per_game`, `Three_Year_Ints_per_game`, `Three_Year_Rush_per_game`).\
-   Join with `player_clean` to attach demographics and write the result to `4_cleaned_nfl_3year.csv`.

```{r message = FALSE, warning = FALSE}
nfl_three_year <- read_csv(here("nfl", "data", "cleaned", "4_cleaned_nfl_3year.csv"),
                      show_col_types = FALSE)
colnames(nfl_three_year)
```

We begin with raw NFL data, construct a cleaned career-level QB dataset, derive early-career summaries (debut and first three years), and finally combine them into the all-in-one analysis file `5_ALL_DATA.csv` documented above.
